E Coli Project Paper Incredibly Rough Draft

### 2013-10-07 paper/paper.tex

A significant amount of work has been done to develop protein reaction
and diffusion models that exhibit accurate macroscopic dynamics of the
MinD protein system. Early models involved free proteins that affect
eachothers' rates of diffusion and membrane attachement but do not
combine into compound states\cite{meinhardt2001pattern} .  In 2003
Huang \cite{huang2003dynamic} improved upon this work with a simple
model based on MinD-MinE combination, ATPase hydrolysis, and MinD
membrane attechent that when simulated exhibits accurate MinD
oscillations in cylindrical cells\cite{huang2003dynamic}. In this
model cytoplasmic MinD is more likely to attach to the membrane when
MinD is already clustered there (following observed non-linear
attachement of minD on the cell membrane), and is stationary once
attached.  A number of studies have used an approach similar in that
they do not rely on the ability of MinD to move along the walls and
cluster ~\cite{kruse2007experimentalist} \cite{meinhardt2001pattern}
\cite{drew2005polymerization} \cite{fange2006noise}
\cite{kerr2006division}, while studies have been made as well of
models which rely on MinD mobility and attraction on the cell
membrane\cite{kruse2002dynamic}\cite{howard2005cellular}.  Variations
of the Huang 2003 model that stochasticly account for variations of
molecular interaction \cite{fange2006noise} and as well monte-carlo
simulations that implement stochastic version of Huang's mean filed
reaction rates confirm the major results obtained by Huang's model and
more successfully predict experimentally observed oscillations in
round cell phenotypes.\cite{drew2005polymerization}
\cite{fange2006noise} \cite{huang2004min} Biochemical models of
broader scope have also been used to study the MinD system and show
consistent results.\cite{arjunan2010new}.  However, in general the
results of the stochastic and monte-carlo simulations are similar to
those given by Huang's mean field results.  They both show the
characteristic oscillations in the pill shaped cells.  We do not see
any compelling reason to avoid using this simple and successful model.

@@

@@

Figure \ref{total-oscillation-plot} shows total protein
results for pill shaped cell of length $5\mu$ and radius $.5\mu$.  The
plot shows both the transerence between different stages of protein
compound and movement within the cell.  Starting at the beginning of a
period, we can see that first there is a spike on the left of MinD-ATP
attached to the wall that is accompanied by a growing number of
cytoplasmic MinD-ATP and MinD-ADP that are entering the left section,
followed by a high peak in MinD-MinE-ATP.  This peak gives way through
the middle to peaks on the right side of the cell, the first being a
small bump in cytoplasmic MinD-ATP that dissapears into a much larger
spike in wall attached MinD-ATP, which is followed by a peak in wall
attached MinD-MinE-ATP.  The pattern during each period clearly shows
the cycle of proteins rections, as well as the very regular nature of
the pole to pole protein oscillations.

@@

>>>>

A significant amount of work has been done to develop protein reaction
and diffusion models that exhibit accurate macroscopic dynamics of the
MinD protein system. Early models involved free proteins that affect
eachothers' rates of diffusion and membrane attachement but do not
combine into compound states\cite{meinhardt2001pattern} .  In 2003
Huang \cite{huang2003dynamic} improved upon this work with a simple
model based on MinD-MinE combination, ATPase hydrolysis, and MinD
membrane attechent that when simulated exhibits accurate MinD
oscillations in cylindrical cells\cite{huang2003dynamic}. In this
model cytoplasmic MinD is more likely to attach to the membrane when
MinD is already clustered there (following observed non-linear
attachement of minD on the cell membrane), and is stationary once
attached.  A number of studies have used an approach similar in that
they do not rely on the ability of MinD to move along the walls and
cluster ~\cite{kruse2007experimentalist} \cite{meinhardt2001pattern}
\cite{drew2005polymerization} \cite{fange2006noise}
\cite{kerr2006division}, while studies have been made as well of
models which rely on MinD mobility and attraction on the cell
membrane\cite{kruse2002dynamic}\cite{howard2005cellular}.  Variations
of the Huang 2003 model that stochasticly account for variations of
molecular interaction \cite{fange2006noise} and as well monte-carlo
simulations that implement stochastic version of Huang's mean filed
reaction rates confirm the major results obtained by Huang's model and
more successfully predict experimentally observed oscillations in
round cell phenotypes.\cite{drew2005polymerization}
\cite{fange2006noise} \cite{huang2004min} Biochemical models of
broader scope have also been used to study the MinD system and show
consistent results.\cite{arjunan2010new}.  However, in general the
results of the stochastic and monte-carlo simulations are similar to
those given by Huang's mean field results.

@@

Our simplest shape is the naturally occuring pill shape.  We peice this shape
together as two hemispherical endcaps attached on either end of a
cylinder.  This shape follows the early simulations of Huang et al but
differs in that we have added the end caps for a more natural shape,
expecting similar results.  We test cells of radius $.5\mu$ and of
lengths $2\mu$, $3\mu$, and $5\mu$ between both tips of the end
caps. In all these models we observe quick establishement (within a
few periods) of extremely regular oscillatory behavoir that lasts for
any length of time, with periods that closely agree with the results
of Huang et al.  The following table shows the periods of oscillations
at different cell lengths.

%% \begin{figure*}
%%   \includegraphics[width=\textwidth]{../data/shape-p/plots/image-plot_E--p-200-50-0-0-1500}
%%   \caption{Frames from an animation showing a series of contour plots
%%     of the concentration of proteins in different stages over time.
%%     The density of proteins are integrated along the axis
%%     normal to the page. The frames occur 2.5 seconds apart.}
%%   \label{total-oscillation-plot}
%% \end{figure*}

This simple model is a good starting point for observing in
detail the dynamic interaction between the different stages of protein
that lead to their qualitatively distinct behavoir. Figure (Figure
image) shows the process in a series of frames taken from simulation
animations.  The causal chain is cyclical so we can begin our
observations at any stage in the cycle.  We look first at MinD:ATP
floating freely in the cytoplasm.  The protein will not react with any
other floating protein, and it will not spontenously shift back into
the MinD:ADP stage of the cycle.  It will diffuse until it attaches to
a cell membrane.  When compared to a protein that is in the middle
cylindrical section of the cell, a protein that is in an end cap
section is more likely to bump into a wall in a given number of
diffusive steps, since there is a greater cross section of possible
movement that is covered by a nearby wall.  All else being equal,
proteins will collect with a higher rate on the caps.  This effect is
made more pronounced by the MinD:ATP tendency to attract and cluster.
When adjacent to the membrane, the rate of binding to the membrane is
porportional to the amount of MinD already bound there.  Initially,
then, there is a strong tendency of MinD to cluster in the polar
regions of the cell.

The next observed mechanism is the formation of MinE 'rings' on the
membranes of the polar caps.  As the freely
diffusing MinE enters the cap portion it begins to encounter
concentrated membrane bound MinD.  As it bumps into the walls here it
binds to this MinD and remians for a time on the walls in compound
form.  When it is released, it can diffuse naturally outward or inward
towards the cell.  If it diffuses inward, there is less cross
sectional wall for it to bump into and also less concentrated MinD:ATP
to react with.  If it diffuses outward, the curvature of the cell wall
cuases there to be a greater chance that it will bump into the cell
membrane (for the same reason outlined for the MinD:ATP above) and if
there is MinD:ATP bound there, there is a chance that it will bind to
that protein and react again.  If it does diffuse in this directin, it
will likely react with MinD:ATP before getting very far into the
encap.  Thus after dettaching from the membrane at the edge of the
MinD:ATP encap cluster, the MinE is most likely to travel a small
distance further into the polar region and then attach and react again
with MinD:ATP protein there.  This process results in the observed
qualitative 'encroaching MinE ring' behavoir that released MinD:ADP
back into the cytoplasm.  This MinD:ADP will diffuse for a time before
spontanosly uindergowing hydrolysis \fixme{is this right?} and
reaching the MinD:ATP state that can bind to the membrane again.
There is a strong chance that it has by this time diffused out of the
polar region and is on its way to the opposite polar region of the
cell.  If it hasn't then it will reattach to the MinD:ATP on the cell
wall deeper into the endcap and will eventually be converged upon by
the MinE ring, only to be released again in MinD:ADP form.

Overall this process results in an extremely robust oscillatory
behavoir of polar selection and oscillation in a wide variety of cell
shapes.

@@

Figure \ref{total-oscillation-plot} is a plot of the total fraction of
MinD found in each stage of the cycle in each of three sections of the
cell (we have split the pill into two polar sections and one center
section) over time.  One complete oscillation period is shown, in
which one of the poles begins and ends at a maximum density
concentration and the other a minimum denisty concentration.  Looking
at the evolution of the latter, we see an initial high concentration
of cytoplasmic MinD:ATP, as the this protein diffuses into the polar
region, followed by a dip in this and large spike in membrane bound
MinD:ATP, as the protein clusters onto the polar cell wall, followed
by a spike in the memebrane bound MinE:MinD:ATP stage of protein,
together with a slight rise in cytoplasmic MinD:ADP as the MinD
dettaches from the wall.

results for pill shaped cell of length
$5\mu$ and radius $.5\mu$.  The plot shows both the transerence
between different stages of protein compound and movement within the
cell.  Starting at the beginning of a period, we can see that first
there is a spike on the left of MinD-ATP attached to the wall that is
accompanied by a growing number of cytoplasmic MinD-ATP and MinD-ADP
that are entering the left section, followed by a high peak in
MinD-MinE-ATP.  This peak gives way through the middle to peaks on the
right side of the cell, the first being a small bump in cytoplasmic
MinD-ATP that dissapears into a much larger spike in wall attached
MinD-ATP, which is followed by a peak in wall attached MinD-MinE-ATP.
The pattern during each period clearly shows the cycle of proteins
rections, as well as the very regular nature of the pole to pole
protein oscillations.

@@

%% \begin{center}
%%   \begin{figure}
%%     {\includegraphics[width=\paperwidth]{../data/shape-randst/plots/image-plot_E--randst-25-602-602-9900-1500}}
%%     \caption{Attempt at adding image plot to paper.}
%%     \label{image-plot}
%%   \end{figure}
%% \end{center}
